Certified Symbolic Transducer with the Application of String Solver

### 2025-02-17 Paper/main.tex
polish: papers

We first present a pure mathematical form of FTs \cite{VeanesHLMB12Transducer}, which is independent of Isabelle's formalization. 
The central concept is the label theory, which characterizes the process by which output labels are derived from input labels.

@@

A lambda term is defined as $\lambda x. t$ of the type $\tau_1 \rightarrow \tau_2$.
When $\tau_2$ is $\mathbb{B}$, this lambda term is a predicate. Assume $\psi$ is a predicate. We write $a\in \llbracket\psi \rrbracket$ if $\psi~a=\top$. For the non-predicate lambda term, we view it as a function, which generates an output element of type $\tau_2$ with the input term of type $\tau_1$.

\begin{definition}[Symbolic Finite Transducer]
\label{def-sft}
   A Symbolic Transducer over $\tau_1\rightarrow \tau_2$ is a quadruple $(\mathcal{Q}, \Delta, \mathcal{I}, \mathcal{F})$, where 
   \begin{itemize}
   \item $\mathcal{Q}$ is a finite set of states,
   \item $\mathcal{I}\subseteq \mathcal{Q}$ is the set of initial states,
   \item $\mathcal{F} \subseteq\mathcal{Q}$ is the set of accepting states,
   \item $\Delta$ is the set of transition relations. Each element in $\Delta$ is of the form $(q, \phi, f, q') (\text{or written in } q\xrightarrow{\phi, f} q')$, where $q$ and $q'$ are the states in $\mathcal{Q}$.
   $\phi$ is a predicate of type $\tau_1\rightarrow \mathbb{B}$.
   $f$ is a lambda term of the type $\tau_1\rightarrow \tau_2$. We call it output function in the following.
   \end{itemize}

For each transition $q\xrightarrow{\phi, f} q'$, if there exists an element $a\in \llbracket \phi \rrbracket$, where $a$ is called input,  then the application $f~a$ is the output.

@@

SFTs accept an input word and generate an output. This can be defined by runs of SFTs.
An SFT run $\sigma$ is a sequence of $(q_0, \psi_0, f_0, q_1),(q_1, \psi_1, f_1, q_2),\ldots, (q_{n-1}, \psi_{n-1}, q_n)$ such that $q_0\in \mathcal{I}_T$ and $(q_i, 
\psi_i, f_i, q_{i+1}), 0 \leq i \leq n-1$ is a transition in $\Delta_T$.
$\sigma$ is an accepting run when $q_n$ is an accepting state.
For a word $w = a_0,\ldots, a_{n-1}$, it is accepted by $\sigma$ iff $a_i \in \llbracket\psi_i \rrbracket$ for $0 \leq i \leq n - 1$. Run $\sigma$ will generate an output sequence if $w$ is accepted by it, which is $w'= (f_0~a_0), \ldots (f_{n_1}~a_{n_1})$. We denote this operation as $\mathcal{N}~w = w'$.

\subsection{The Isabelle/HOL Formalization of SFT}

@@

\section{The Key Operation of NFTs}
\label{sec-form-fts}

@@

In order to prove this abstract definition of product operation's correctness, we firstly formalize the outputs of an NFT regarding the input of an NFA. Figure \ref{fig-def-output} has two definitions: (1) \texttt{outputE} extracts the output (second element) from a path in the NFT. (2) \texttt{outputL} defines the set of all outputs of an NFT $\mathcal{T}$, when the input is in the  language $\mathcal{A}$.

>>>>

We first present a pure mathematical form of SFTs \cite{VeanesHLMB12Transducer}, which is independent of Isabelle's formalization. 
The central concept is the label theory, which characterizes the process by which output labels are derived from input labels.

@@

A lambda term is defined as $\lambda x. t$ of the type $\tau_1 \rightarrow \tau_2$.
When $\tau_2$ is $\mathbb{B}$, this lambda term is a predicate. Assume $\psi$ is a predicate. We write $a\in \llbracket\psi \rrbracket$ if $\psi~a=\top$. For the non-predicate lambda term, we view it as a function, which generates an output element of type $\tau_2$ with the input term of type $\tau_1$. 
With these notations and the above definitions, we can define SFTs as follows.

\begin{definition}[Symbolic Finite Transducer]
\label{def-sft}
   A Symbolic Finite Transducer over $\tau_1\rightarrow \tau_2$ is a quadruple $\mathcal{T} = (\mathcal{Q}, \Delta, \mathcal{I}, \mathcal{F})$, where 
   \begin{itemize}
   \item $\mathcal{Q}$ is a finite set of states,
   \item $\mathcal{I}\subseteq \mathcal{Q}$ is the set of initial states,
   \item $\mathcal{F} \subseteq\mathcal{Q}$ is the set of accepting states,
   \item $\Delta$ is the set of transition relations. Each element in $\Delta$ is of the form $(q, \psi, f, q') \text{ or}$ $\text{written in }q\xrightarrow{\psi, f} q'$, where $q$ and $q'$ are the states in $\mathcal{Q}$.
   $\psi$ is a predicate of type $\tau_1\rightarrow \mathbb{B}$.
   $f$ is a lambda term of the type $\tau_1\rightarrow \tau_2$. $f$ is called "output function".
   \end{itemize}

For each transition $q\xrightarrow{\phi, f} q'$, if there exists an element $a\in \llbracket \phi \rrbracket$, where $a$ is called an input,  then the application $(f~a)$ is the output.

@@

SFTs accept an input word and generate an output. This can be defined by \emph{runs} of SFTs.
An SFT run $\sigma$ is a sequence $(q_0, \psi_0, f_0, q_1),(q_1, \psi_1, f_1, q_2),\ldots, (q_{n-1}, \psi_{n-1}, q_n)$ such that $q_0\in \mathcal{I}$ and $(q_i, 
\psi_i, f_i, q_{i+1}), 0 \leq i \leq n-1$ is a transition in $\Delta$.
$\sigma$ is an accepting run when $q_n$ is an accepting state.

For a word $w = a_0,\ldots, a_{n-1}$, it is accepted by $\sigma$ iff $a_i \in \llbracket\psi_i \rrbracket$ for $0 \leq i \leq n - 1$. Run $\sigma$ will generate an output sequence if $w$ is accepted by it, which is $w'= (f_0~a_0), \ldots (f_{n_1}~a_{n_1})$. We call in the sequent:
\begin{itemize}
\item  $(a_0,(f_0~a_0)), \ldots, (a_{n-1},(f_{n-1}~a_{n-1}))$ a \emph{path},
\item $a_0,\ldots,a_{n-1}$ the \emph{input} of the path and $(f_0~a_0), \ldots, (f_{n-1}~a_{n-1})$ the \emph{output} of the path.
\end{itemize}

If $a_0,\ldots,a_{n-1}$ is accepted by an accepting run in $\mathcal{T}$, we say that the path is an accepting path of $\mathcal{T}$.
For a path $\pi$, we denote the input of $\pi$ as $in(\pi)$ and the output of $\pi$ as $out(\pi)$.
%
Given an SFT $\mathcal{T}$ and a word $w$, we define the operation product of $\mathcal{T}$ and $w$ (denoted as $\mathcal{T}\times\{w\}$) as the set of outputs generated by $\mathcal{T}$ with the input $w$. More precisely, 
\[
\mathcal{T}\times\{w\} = \{w'\mid \exists \pi.~\pi \text{ is an accepting path }\land in(\pi) = w \land out(\pi) = w'\}.
\]

To make the operation \emph{product} more general, we extend the operation to an SFT and a set of intput words represented by a regular language, which can be denoted as an NFA. More precisely,
\[
\mathcal{T}\times\{\mathcal{A}\} = \{w'\mid w.~w\in \mathcal{L}(\mathcal{A})\land w' = \mathcal{T}\times\{w\} \}, \text{where $\mathcal{L}(\mathcal{A})$ denotes the language of $\mathcal{A}$}.
\]

\begin{figure}[hbt!]
  \centering
  \begin{tikzpicture}[shorten >=1pt, node distance=3.5cm, on grid, auto]
    \node[state, initial] (q_0)   {$q_0$}; 
    \node[state] (q_1) [right=of q_0] {$q_1$}; 
    \node[state, accepting] (q_2) [right=of q_1] {$q_2$};

    \path[->] 
    (q_0) edge node {[$'\texttt{a}'-'\texttt{z}'$],$f_1$} (q_1)
    (q_1) edge node {[$'\texttt{A}'-'\texttt{Z}'$],$f_2$} (q_2)
    (q_1) edge[loop above] node {[$'\texttt{a}'-'\texttt{z}'$],$f_1$} (q_1)
    (q_2) edge[loop above] node {[$'\texttt{A}'-'\texttt{Z}'$],$f_2$} (q_2);
 \end{tikzpicture}
    \caption{The formalization of $FTs$ in Isabelle/HOL}
    \label{fig-example-ft}
    \end{figure}   
We now illustrate SFTs with an example in Figure \ref{fig-example-ft}. It accepts strings of the regular expression $/['\texttt{a}'-'\texttt{z}']+['\texttt{A}'-'\texttt{Z}']+/$. 
It uses intervals $[i-j]$ as the input label, which is a predicate that returns true if the input letter is between $i$ and $j$.
The function $f_1=\lambda x.~ \texttt{toUpper}(x)$ maps each lowercase letter to uppercase letter, and $f_2=\lambda x.~\texttt{toLower}(x)$ maps each uppercase letter to lowercase letter.
Give an string "$\texttt{bigSMALL}$", the SFT will generate an output "$\texttt{BIGsmall}$".

\subsection{The Isabelle/HOL Formalization of SFTs}

@@

\section{The Product Operation of NFTs}
\label{sec-form-fts}

@@

To establish the correctness of the abstract definition of the product operation, we first formalize the outputs of a Non-deterministic Finite Transducer (NFT) with respect to the input of a Non-deterministic Finite Automaton (NFA). Figure \ref{fig-def-output} presents two key definitions: (1) \texttt{outputE}, which extracts the sequence of outputs $o_1, \ldots, o_n$ from a given NFT path $(i_1, o_1), \ldots, (i_n, o_n)$; and (2) \texttt{outputL}, which represents the set of all possible outputs generated by an NFT $\mathcal{T}$ for the input language defined by an NFA $\mathcal{A}$.
